---
title: "ESM 244 Lab 1"
author: "Kira Archipov"
date: "2023-01-13"
output: html_document
---

```{r setup, include=TRUE, echo = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(sf)
#stands for simple features
library(tmap)
#good for visualizing map data

```

Workflow best practices: 
commit pull push

In terminal: 
ls - tells you all the folders within current repository 
ls -a - includes .git folder within project 

Environment over on top right - tells you which variables are in R memory 
read_csv is improved version of read.csv within tidyverse, faster and does a couple more things than read.csv

### Read in the data 
```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'))

top_5_status <- sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n()) %>% 
  slice_max(tree_count, n = 5) %>% #find values that are max and just keep n amount 
  arrange(-tree_count)

ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count),
                                y = tree_count)) +
  geom_col() + 
  labs(y = 'Tree Count', x = 'Legal Status') + coord_flip() + theme_minimal()
```

### Playing with filtering - keep obs where legal status is "permitted sites" and aretaker is MTA
```{r}
#sf_trees$legal_status %>% unique()
# gives you list of all unique values of legal status column 

permitted_mta <- sf_trees %>% filter(legal_status =="Permitted Site" & caretaker == "MTA")

```

### Keep blackwood acacia trees , only legal status, date, lat and long columns 
```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, latitude, longitude)


#Make a plot of SF acacia trees

ggplot(data = blackwood_acacia, aes(x = longitude, y = latitude)) + 
  geom_point()
```

## Use tidyR::separate()
use it to separate one column into multiple 

### use tidyR::unite() to rejoin 
```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')

#reverse we want to combine two (or multiple) columns into one 

sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id:species, sep = '_NEW_')
```

### Make some actual maps !
```{r}
blackwood_acacia_sf <- blackwood_acacia %>% 
  drop_na(latitude, longitude) %>% 
  st_as_sf(coords = c('longitude', 'latitude'))

st_crs(blackwood_acacia_sf) <- 4326
#numeric code that sets basic lat/long system 

ggplot(data = blackwood_acacia_sf) +
  geom_sf(color = 'dark green') +
  theme_minimal()
#geom_sf automatically looks for geometry collumn 
```


### read in SF street data 
```{r}
sf_map_sf <- read_sf(here('data', 'sf_map', 'tl_2017_06075_roads.shp')) %>% 
  st_transform(4326)

st_crs(sf_map_sf)

```

### Mappin it out 
```{r}
ggplot() +
  geom_sf(data = sf_map_sf, size = 0.1, color = 'darkgrey') +
  geom_sf(data = blackwood_acacia_sf, color = 'red', size = 0.5) + 
  theme_void() +
  labs(title = "Blackwood acacias in San Fransisco")
```

### INteractive map 
```{r}
tmap_mode('view') 
tm_shape(blackwood_acacia_sf) + 
  tm_dots()
```













